Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-10-29 manuscript_latex/manuscript.tex
Update manuscript.tex

%%Note: the following reference styles support Namedate and Numbered referencing. By default the style follows the most common style. To switch between the options you can add or remove “Numbered” in the optional parenthesis. 
%%The option is available for: sn-basic.bst, sn-vancouver.bst, sn-chicago.bst%

@@

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
%% \usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
%% \usepackage{algorithm}%
%% \usepackage{algorithmicx}%
%% \usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{placeins}
%%%%
\newtheorem*{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}

@@

{\centering \includegraphics[width=1\linewidth]{figure_1}

@@

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in \hyperref[appendix-a]{Appendix A}.
The location- and scale-invariant Box-Cox log-likelihood function is:

@@

{\centering \includegraphics[width=1\linewidth]{figure_2}

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After \citep{Raymaekers2024-zf},
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) \citep{Huber1981-su}. Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the quantile function
\(F_{\alpha}^{-1}(p)\), which incorporates the quantile function of the
symmetric generalised normal distribution derived by Griffin \citep{Gijbels2019-te}:

@@

\subsection{Central normality test}\label{central-normality-test}

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test  \citep{Gijbels2019-te}. In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop a test for central normality.

@@

Let the residual error for each element of $\mathbf{X}$ be $r_i =\left| \frac{x_i - \mu_M}{\sigma_M} - F^{-1}_{\mathcal{N}}(p_i) \right|$, 
with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of $\mathbf{X}$ \citep{Huber1981-su}, 
and $F^{-1}_{\mathcal{N}}$ the quantile function of the normal distribution $\mathcal{N}(0, 1)$.

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation \citep{Powell2009-zb} 
through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
\citep{Zwanenburg2024-kq} (version 1.0.1). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

{\centering \includegraphics[width=1\linewidth]{figure_3}

@@

{\centering \includegraphics[width=1\linewidth]{figure_4}

@@

\caption{Critical test statistic $\tau_{\alpha = 0.05, n, \kappa}$ of the (empirical) central normality test as function of $n$ for several values of central portion $\kappa$. The critical test statistics for central normality test are determined using fully normal data, whereas the statistics for the empirical variant are determined using centrally normal data, i.e. with fully normal data where 10\% of elements are replaced by outliers. emp: empirical}\label{fig:empirical-central-normality-kappa}
\end{figure}

To assess type I error rates for the (empirical) central normality test
and compare these to the Shapiro-Wilk test for normality, we randomly
drew another 1000 sequences from a standard normal distribution
(\(\mathcal{N}(0,1)\) for each
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}\).
A second dataset was created by replacing elements with randomly drawn
outliers, as described above. For each sequence, the p-value for the
respective test was used to reject the null hypothesis that the sequence
is (centrally) normal. The null hypothesis was rejected if
\(p \leq 0.05\). The type I error rate was determined by computing the
fraction of centrally normal sequences rejected this way. Figure
\ref{fig:empirical-central-normality-error_rate} shows that the
empirical central normality test had the expected type I error rate for
centrally normal data (with tails of the distribution not being normally
distributed), whereas the central normality test and Shapiro-Wilk test
overestimated type I error rates.

@@

{\centering \includegraphics[width=1\linewidth]{figure_5}

@@

\caption{Observed type I error rates for (central) normality tests in the presence of data derived from normal distributions with and without outliers. The empirical central normality (ECN) test uses $\kappa = 0.80$. CN: central normality}\label{fig:empirical-central-normality-error_rate}
\end{figure}

In Figure \ref{fig:empirical-central-normality-examples} we apply the
empirical central normality test to assess central normality of features
that are composed of a mixture of samples drawn from two normal
distributions (\(n = 100\) each). With increased separation of the
underlying normal distributions, the probability of the feature being
centrally normal decreases, as expected.

@@

{\centering \includegraphics[width=1\linewidth]{figure_6}

@@

\caption{Bi-modal distributions and empirical central normality test results. The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) drawn from normal distributions that are offset by a distance $d$. We use the empirical centrally normality test to compute the probability for the hypothesis that the distribution is centrally normal. As may be observed, with increasing offset $d$ the probability that the feature is centrally normal decreases. Quantile-quantile plots are drawn below each distribution.}\label{fig:empirical-central-normality-examples}
\end{figure}

@@

To determine the weighting function parameters for each of the nine
combinations, \(m_d=500\) sequences \(\mathbf{X}_i\)
(\(i \in \{1, 2, \ldots, m_d\}\)) were randomly drawn from randomly
parametrised asymmetric generalised normal distributions. Each
distribution was parametrised with a random skewness parameter
\(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
\(\beta \sim U\left(1.00, 5.00 \right)\). Location and scale parameters
were set as \(\mu = 0\) and \(\sigma = 1\), respectively. To form each
sequence \(\mathbf{X}_i\), \(n = \lceil 10^\gamma \rceil\) instances
were then randomly drawn, with
\(\gamma \sim U\left(\log_{10}50, 3\right)\), resulting in a set of
sequences with between \(50\) and \(1000\) elements each.

Outlier values were then drawn to randomly replace 10 percent of the
elements of \(\mathbf{X}_i\). Outlier values were set according to \citet{Tukey1977-xm},
as follows. Let \(x^{*} \sim U\left(-2, 2\right)\).
Then the corresponding outlier value was:

\begin{equation}
x_{out} =
\begin{cases}
Q_1 - \left(1.5 - x^{*} \right) \text{IQR} & \text{if } x^{*} < 0 \\
Q_3 + \left(1.5 + x^{*} \right) \text{IQR} & \text{if } x^{*} \geq 0
\end{cases}
\end{equation}

\(Q_1\), \(Q_3\) and \(\text{IQR}\) are the first quartile, third
quartile and interquartile range of \(\mathbf{X}_i\), respectively.
Outlier values randomly replaced elements in \(\mathbf{X}_i\).

@@

\begin{table}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Box-Cox power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution.
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. 
The \textit{optimal} column shows the optimal parameter values.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-box-cox}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 49.6 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.80 & ---  & ---       & ---  & 38.5 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.01 & 1.00 & $(0, 1]$  & 0.86 & 43.7 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.00 & 1.00 & $(0, 1]$  & 0.90 & 42.7 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.39 & ---  & ---       & ---  & 52.2 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 2.39 & 2.40 & $(0, 10]$ & 4.92 & 52.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 1.07 & 3.63 & $(0, 10]$ & 3.43 & 57.4 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.83 & ---  & ---       & ---  & 53.1 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.80 & 0.81 & $(0, 10]$ & 0.84 & 54.4 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 0.76 & 0.77 & $(0, 10]$ & 0.76 & 53.2 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Yeo-Johnson power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. 
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-yeo-johnson}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 42.1 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.78 & ---  & ---       & ---  & 35.1 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.20 & 0.95 & $(0, 1]$  & 1.00 & 34.0 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.54 & 0.95 & $(0, 1]$  & 1.00 & 32.6 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.32 & ---  & ---       & ---  & 43.0 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 1.28 & 1.96 & $(0, 10]$ & 3.43 & 49.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 0.41 & 1.96 & $(0, 10]$ & 3.75 & 51.3 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.92 & ---  & ---       & ---  & 65.4 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.87 & 1.00 & $(0, 10]$ & 0.87 & 66.0 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 1.06 & 1.00 & $(0, 10]$ & 1.07 & 67.1 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of sequence
  \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / \text{IQR}\left(\mathbf{X}\right)\),
  with \(\text{IQR}\) representing the interquartile range of
  \(\mathbf{X}\).
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 10000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulations_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in \hyperref[appendix-e]{Appendix E}.

@@

{\centering \includegraphics[width=1\linewidth]{figure_7}

@@

{\centering \includegraphics[width=1\linewidth]{figure_8}

@@

{\centering \includegraphics[width=1\linewidth]{figure_9}

>>>>

%%Note: the following reference styles support Namedate and Numbered referencing. By default the style follows the most common style. To switch between the options you can add or remove ?Numbered? in the optional parenthesis.
%%The option is available for: sn-basic.bst, sn-vancouver.bst, sn-chicago.bst%

@@

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
%% \usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
%% \usepackage{algorithm}%
%% \usepackage{algorithmicx}%
%% \usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{placeins}
%%%%
\newtheorem*{definition}{Definition}
\newtheorem*{corollary}{Corollary}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}

@@

{\centering \includegraphics{manuscript_files/figure-latex/decreased-normality-1}

@@

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in Appendix A.
The location- and scale-invariant Box-Cox log-likelihood function is:

@@

{\centering \includegraphics{manuscript_files/figure-latex/weighting-functions-1}

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After
  \citet{Raymaekers2024-zf},
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) \citep{Huber1981-su}. Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the quantile function
\(F_{\alpha}^{-1}(p)\), which incorporates the quantile function of the
symmetric generalised normal distribution derived by Griffin
\citep{Griffin2018-bf}:

@@

\subsection{Empirical central normality
test}\label{empirical-central-normality-test}

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test \citep{Shapiro1965-zd}. In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality.

@@

Let the residual error for each element of $\mathbf{X}$ be $r_i =\left| \frac{x_i - \mu_M}{\sigma_M} - F^{-1}_{\mathcal{N}}(p_i) \right|$,
with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of $\mathbf{X}$ [@Huber1981-su],
and $F^{-1}_{\mathcal{N}}$ the quantile function of the normal distribution $\mathcal{N}(0, 1)$.

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation
\citep{Powell2009-zb} through maximum likelihood estimation. The
required algorithms were implemented in the \texttt{power.transform} R
software package \citep{Zwanenburg2024-kq} (version 1.0.2). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

{\centering \includegraphics{manuscript_files/figure-latex/shifted-distributions-1}

@@

{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-combined-1}

@@

\caption{Critical test statistic $\tau_{\alpha = 0.05, n, \kappa}$ of the empirical central normality test as function of $n$ for several values of central portion $\kappa$ (left) and several values of significance level $\alpha$ (right).}\label{fig:empirical-central-normality-combined}
\end{figure}

To assess type I error rates for the empirical central normality test
and compare these to the Shapiro-Wilk test for normality, we randomly
drew another 10000 sequences from a central normal distribution for each
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}\).

Each sequence was created as follows: first \(n\) values are sampled
from \(\mathcal{N}(0, 1)\). Then, the elements outside the central
portion of the sequence
(i.e.~\(\mathbf{X} \setminus \mathbf{X}_{\text{c}}\)) are replaced.
Elements with \(p_i < \frac{1-\kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))\), and
elements with \(p_i > \frac{1 + \kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)\).

For each sequence, the p-value for the respective test was used to
reject the null hypothesis that sequences were derived from a population
with a (central) normal distribution. The null hypothesis was rejected
if \(p \leq 0.05\). The type I error rate was determined by computing
the fraction of sequences rejected this way. Figure
\ref{fig:empirical-central-normality-error_rate} shows that when the
central portion \(\kappa\) of the central normal distribution is equal
or greater than the \(\kappa\) parameter for the empirical central
normality test, the tests are accurate. If \(\kappa\) of the empirical
central normality test exceeds that the central portion \(\kappa\) for
the underlying distribution, tests are conservative. The Shapiro-Wilk
test behaves similar to the empirical central normality test with
\(\kappa = 1.0\), which is expected. For \(n < 10\), the empirical
central normality test can be optimistic.

@@

{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-error_rate-1}

@@

\caption{Observed type I error rates for central normality tests in the presence of data sampled from central normal distributions with $\kappa \in \left\{0.6, 0.8, 1.0\right\}$. Three empirical central normality (ECN) tests were performed, parameterised with same value for the central portion $\kappa$. Additionally, the Shapiro-Wilk test was assessed. The dashed line indicates the expected type I error rate for rejecting the null hypothesis at significance level $\alpha = 0.05$. ECN: empirical central normality}\label{fig:empirical-central-normality-error_rate}
\end{figure}

In Figure \ref{fig:empirical-central-normality-examples} we apply the
empirical central normality test (with \(\kappa = 0.80\)) to assess
central normality of features that are composed of a mixture of samples
drawn from two normal distributions (\(n = 100\) each). With increased
separation of the underlying normal distributions, the test's p-value
decreases, as expected.

@@

{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-examples-1}

@@

\caption{Bi-modal distributions and empirical central normality test results. The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) drawn from normal distributions that are offset by a distance $d$. The empirical central normality test ($\kappa = 0.80$) is used to test the hypothesis that the mixture is sampled from a central normal distribution. As may be observed, with increasing offset $d$ the test's p-value decreases. Quantile-quantile plots are drawn below each distribution.}\label{fig:empirical-central-normality-examples}
\end{figure}

We use \(\kappa = 0.80\) for the empirical central normality tests in
the remainder of the manuscript, unless noted otherwise. Critical
statistic values of the empirical central normality test for
\(\kappa = 0.80\) are shown in Table
\ref{tab:empirical-central-normality-critical-statistic}.

@@

To determine the weighting function parameters for each of the nine
combinations, \(m_d=500\) sequences \(\mathbf{X}_i\)
(\(i \in \{1, 2, \ldots, m_d\}\)) were randomly drawn from randomly
parametrised asymmetric generalised normal distributions. Each
distribution was parametrised with a random skewness parameter
\(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
\(\beta \sim U\left(1.00, 5.00 \right)\). Location and scale parameters
were set as \(\mu = 0\) and \(\sigma = 1\), respectively. To form each
sequence \(\mathbf{X}_i\), \(n = \lceil 10^\gamma \rceil\) instances
were then randomly drawn, with
\(\gamma \sim U\left(\log_{10}50, 3\right)\), resulting in a set of
sequences with between \(50\) and \(1000\) elements each. Outlier values
were then drawn to randomly replace 10 percent of the elements of
\(\mathbf{X}_i\). These values were set according to
\citet{Tukey1977-xm}, as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

\begin{equation}
\label{eqn:outlier_rate}
x_{out} =
\begin{cases}
Q_1 - \left(1.5 - x^{*} \right) \text{IQR} & \text{if } x^{*} < 0 \\
Q_3 + \left(1.5 + x^{*} \right) \text{IQR} & \text{if } x^{*} \geq 0
\end{cases}
\end{equation}

Here, \(Q_1\), \(Q_3\) and \(\text{IQR}\) are the first quartile, third
quartile and interquartile range of \(\mathbf{X}_i\), respectively.

@@

\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Box-Cox power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution.
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column.
The \textit{optimal} column shows the optimal parameter values.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-box-cox}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 49.6 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.80 & ---  & ---       & ---  & 38.5 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.01 & 1.00 & $(0, 1]$  & 0.86 & 43.7 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.00 & 1.00 & $(0, 1]$  & 0.90 & 42.7 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.39 & ---  & ---       & ---  & 52.2 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 2.39 & 2.40 & $(0, 10]$ & 4.92 & 52.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 1.07 & 3.63 & $(0, 10]$ & 3.43 & 57.4 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.83 & ---  & ---       & ---  & 53.1 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.80 & 0.81 & $(0, 10]$ & 0.84 & 54.4 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 0.76 & 0.77 & $(0, 10]$ & 0.76 & 53.2 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Yeo-Johnson power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution.
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-yeo-johnson}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 42.1 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.78 & ---  & ---       & ---  & 35.1 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.20 & 0.95 & $(0, 1]$  & 1.00 & 34.0 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.54 & 0.95 & $(0, 1]$  & 1.00 & 32.6 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.32 & ---  & ---       & ---  & 43.0 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 1.28 & 1.96 & $(0, 10]$ & 3.43 & 49.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 0.41 & 1.96 & $(0, 10]$ & 3.75 & 51.3 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.92 & ---  & ---       & ---  & 65.4 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.87 & 1.00 & $(0, 10]$ & 0.87 & 66.0 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 1.06 & 1.00 & $(0, 10]$ & 1.07 & 67.1 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of sequence
  \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / Q_n\left(\mathbf{X}\right)\),
  with \(Q_n\) representing the robust \(Q_n\) scale estimator of
  \(\mathbf{X}\) \citep{Croux1992-zz, Rousseeuw1993-it}.
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 100000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulation_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in Appendix D.

@@

{\centering \includegraphics{manuscript_files/figure-latex/experimental-results-invariance-1}

@@

{\centering \includegraphics{manuscript_files/figure-latex/experimental-results-outlier-robustness-1}

@@

{\centering \includegraphics{manuscript_files/figure-latex/marginal-effect-plot-1}
